---
title: Genomics-Assisted Breeding
editor: visual
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    code-tools:
      source: true
      toggle: false
      caption: none
---

This is a notebook for...

## Resources

- [breedR Tutorial](https://github.com/famuvie/breedR/wiki/Getting-started)
- [ASRgenomics Manual](https://asreml.kb.vsni.co.uk/wp-content/uploads/sites/3/ASRgenomics_Manual.pdf)

```{r, libraries, message=FALSE}
#| code-summary: Show the code that imports packages.

library(ASRgenomics)     # Molecular and genetic routine companions for ASReml-R, Genomic Selection (GS), and Genome-Wide Association Analyses (GWAS)
library(breedR)          # Statistical methods for forest genetic resources analysts
library(dplyr)           # Enables the %>% operator for command pipes.
library(patchwork)       # Compose figures made of multiple subfigures
library(purrr)           # Functional programming, captures messages and other outputs
library(reactablefmtr)   # Visualize tables
```

## ASRgenomics

### Reading and Filtering a Molecular Dataset

::: {.callout-note collapse="true" appearance="simple" icon="false"}
To illustrate the flow of functions and show some of the capabilities of ASRgenomics we are going to use a real dataset from Loblolly Pine (Pinus taeda L.) published by Resende et al. (2012). The genotypic portion of this dataset contains a total of 655 genotypes and 4,853 SNPs (coded as 0, 1, and 2, and here -9 is used for missing data). This is a subset of the original dataset where some full-sib genomic records have been artificially eliminated for illustration purposes.
:::

```{r}
#| code-summary: Show the code to load genotypes.
#| label: tbl-data-pine655
#| tbl-cap: "Loblolly Pine table of 655 genotypes and 4,853 SNPs (-9 indicates missing data)<br>Only the first 4 genotypes and SNPs are displayed for brevity."
#| tbl-cap-location: top
data(geno.pine655)
reactable(data = geno.pine655[1:4, 1:4])
```

::: {.callout-note collapse="true" appearance="simple" icon="false"}
A reasonable initial step is to filter this molecular dataset, as it might contain large amounts of missing values and/or non-informative markers. This quality control step is critical to facilitate downstream analyses and further calculations (particularly to ensure that the genomic-based matrix G is invertible. For this, the function `qc.filtering()` can be used.

Since the proportion of missing values in this dataset is less that 5%, we will perform a simple imputation (average method) of missing values.
:::

```{r,fig.height = 6}
#| code-summary: Show the code to filter genotypes.
#| label: filter-pine655

# Redirect important messages to a variable
quiet_filtering <- purrr::quietly(qc.filtering)
result <- quiet_filtering(
    M = geno.pine655,
    base = FALSE,
    ref = NULL,
    marker.callrate = 0.2,
    ind.callrate = 0.2,
    maf = 0.05,
    heterozygosity = 0.95,
    Fis = 1,
    impute = TRUE,
    na.string = "-9",
    plots = TRUE
)

# Store the analytical results
M_filter <- result$result
M_clean  <- as.data.frame(M_filter$M.clean)

M_filter$plot.missing.ind +
    M_filter$plot.missing.SNP +
    M_filter$plot.Fis +
    M_filter$plot.heteroz +
    M_filter$plot.maf +
    plot_layout(ncol = 3, nrow=2)
```


```{r, results='asis', echo=FALSE}
#| label: filter-pine655-messages

# Display the messages in a nicer format.
cat(
    '::: {.callout-note collapse="true" appearance="simple" icon="false"}\n',
    paste0("- ", result$messages),
    ':::',
    sep=""
)
```

### Generating a Kinship Matrix

## breedR

### Pedigree

::: {.callout-note collapse="true" appearance="simple" icon="false"}
-   A pedigree table identifies individual trees and their parents.
:::

```{r}
#| code-summary: Show the code to construct the pedigree.
#| label: tbl-pedigree-raw
#| tbl-cap: A pedigree table that is unsorted and incomplete due to unsampled parents.
#| tbl-cap-location: top

# Generate an example pedigree table
set.seed(123)
n.ped <- 5
ped.raw <- matrix(
    sample(30, n.ped*3),
    n.ped, 3,
    dimnames = list(NULL, c('self', 'sire', 'dam'))
)

# Customize the visualization per columns
column_options <- list (
    self = colDef(name = "Tree"),
    sire = colDef(name = "Male"),
    dam  = colDef(name = "Female")
)

# Visualize the pedigree
reactable(data = ped.raw, columns = column_options)
```

::: {.callout-note collapse="true" appearance="simple" icon="false"}
-   The pedigree is sometimes unsorted and incomplete, with missing parent rows.
-   The `breedR` package implements a method to complete, sort and recode the pedigree.
:::

```{r, warning=FALSE}
#| code-summary: Show the code to fix the pedigree.
#| label: tbl-pedigree-fix
#| tbl-cap: A pedigree table that is sorted and complete.
#| tbl-cap-location: top

# Fix any pedigree format issues
ped.fix <- as.data.frame(build_pedigree(1:3, data = ped.raw))

# Visualize the fixed pedigree
reactable(data = ped.fix, columns = column_options, pagination=FALSE)
```

### Additive Genetic Effects

- Additive genetic effect, restricted maximum-likelihood (REML)

```{r, }
test.dat <- data.frame(ped.raw, y = rnorm(n.ped))
res.raw <- remlf90(
    data    = test.dat,
    fixed   = y ~ 1,
    genetic = list(
        id = 'self',
        model = 'add_animal',
        pedigree = ped.raw,
        var.ini = 1
      ),
    var.ini = list(resid = 1),
)
#reactable(data = res.raw)
```
